---
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dpi = 150
)
```

# msaviz

<!-- badges: start -->
[![R-CMD-check](https://github.com/bjmt/msaviz/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/bjmt/msaviz/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`msaviz` is a small R package for visualizing multiple sequence alignments
(MSAs). It reshapes an alignment into a tidy `tibble` and renders it
either as a ggplot2 heatmap or as a per-sequence lollipop chart, with a
`rasterGeom()` wrapper that keeps file sizes and render times sane even
for alignments with millions of cells.

The code in this package was originally written to produce some of the
figures in [Hodgins et al. (2023), *Nature Communications*
14:5475](https://www.nature.com/articles/s41467-023-41174-0), and was later
packaged up into this form.

## Installation

```{r install, eval = FALSE}
# install.packages("remotes")
remotes::install_github("bjmt/msaviz")
```

`msaviz` does not depend on any particular alignment reader. Use either
[seqinr](https://cran.r-project.org/package=seqinr) or
[Biostrings](https://bioconductor.org/packages/Biostrings/) to load your
sequences.

## Quick example

```{r example, fig.width = 7, fig.height = 3}
library(msaviz)

aln_path <- system.file(
  "extdata", "hug_ribosomal.fa.gz", package = "msaviz"
)
aln <- seqinr::read.alignment(aln_path, format = "fasta")

alnDF <- msa2DF(aln)
msaHeatmap(alnDF)
```

By default `msaHeatmap()` colours every cell by whether it matches the
reference sequence (a computed consensus if you don't pass `reference =`).
Switch to per-letter colouring with `column = "Letter"`.

## Per-sequence lollipops

`msaLollipop()` is an alternative view of the same `alnDF`. Each
sequence becomes one row, every non-reference cell is drawn as a
vertical stem with a circular head, and a per-position baseline shows
gaps as breaks. This is the form used for Supplementary Figure 20 of
Hodgins et al. (2023).

```{r lollipop, fig.width = 8, fig.height = 5}
alnDF_ll <- msa2DF(aln, drop.gaps = FALSE, keep.consensus = TRUE)
msaLollipop(alnDF_ll, labels = FALSE)
```

On smaller alignments (or short windows of larger ones), labels are
readable and heads can be colour-mapped to any column:

```{r lollipop-small, fig.width = 8, fig.height = 2.5}
toy <- c(
  seq1 = "ACGTACGTACGTACGTACGT",
  seq2 = "ACGTACGAACGTACGTACGT",
  seq3 = "AAGTACGTACGTGCGTACGT",
  seq4 = "ACGTACGTACGTACGTCCGA"
)
toyDF <- msa2DF(toy, reference = "seq1", drop.gaps = FALSE,
  keep.consensus = TRUE)
msaLollipop(toyDF,
  head.fill.by = "Letter",
  head.fill.colours = msa_palette_DNA)
```

## Saving plots

`saveHeatmap()` is a `ggsave()` wrapper that picks a reasonable size based on
the number of sequences and axis text size:

```{r save, eval = FALSE}
saveHeatmap(msaHeatmap(alnDF), "aln.pdf")
```

## Shell scripts

Two standalone CLIs ship with the package. `inst/scripts/msaHeatmap.R`
renders an alignment heatmap; `inst/scripts/msaLollipop.R` renders the
per-sequence lollipop chart (the form used for Supplementary Figure 20
of Hodgins et al. 2023).

```{r cli-path, eval = FALSE}
system.file("scripts", "msaHeatmap.R",  package = "msaviz")
system.file("scripts", "msaLollipop.R", package = "msaviz")
```

Use them as:

```sh
Rscript msaHeatmap.R  --input aln.fa --output aln.pdf      --column Aln
Rscript msaLollipop.R --input aln.fa --output lollipop.pdf --reference E88
```

See `--help` for all available options on each.

## Composing with PID, groups, and a dendrogram

`composeMSA()` lines the heatmap up (using `patchwork`) with optional
companion plots: a percent-identity track, a hierarchical-clustering
dendrogram, and a left-side group annotation strip.

```{r compose, fig.width = 8, fig.height = 5}
d <- msaDendro(aln)
groups <- setNames(rep(c("Bacteria", "Archaea", "Eukarya"),
                       length.out = length(aln$nam)),
                   aln$nam)
group_cols <- c(Bacteria = "#1f77b4", Archaea = "#ff7f0e",
                Eukarya  = "#2ca02c")

composeMSA(
  heatmap = msaHeatmap(alnDF, row.order = attr(d, "order"),
                       legend.pos = "none"),
  top  = msaPID(alnDF),
  left = d,
  groups = groups[attr(d, "order")],
  group.colours = group_cols,
  heights = c(1, 6),       # PID short, heatmap tall
  widths  = c(0.5, 0.04, 4) # narrow dendrogram, thin group strip, wide heatmap
)
```

## More

The package vignette walks through `Aln` vs `Letter` colouring, custom
palettes, SNP emphasis, the `raster = TRUE/FALSE` trade-off, `posStats()`,
the lollipop view, and the rest of the companion plotters:

```{r vignette, eval = FALSE}
vignette("msaviz")
```
