---
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dpi = 150
)
```

# msaviz

<!-- badges: start -->
[![R-CMD-check](https://github.com/bjmt/msaviz/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/bjmt/msaviz/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`msaviz` is a small R package for visualizing multiple sequence alignments
(MSAs). It reshapes an alignment into a tidy `tibble` and turns it into a
ggplot2 heatmap, with a `rasterGeom()` wrapper that keeps file sizes and
render times sane even for alignments with millions of cells.

## Installation

```{r install, eval = FALSE}
# install.packages("remotes")
remotes::install_github("bjmt/msaviz")
```

`msaviz` does not depend on any particular alignment reader. Use either
[seqinr](https://cran.r-project.org/package=seqinr) or
[Biostrings](https://bioconductor.org/packages/Biostrings/) to load your
sequences.

## Quick example

```{r example, fig.width = 7, fig.height = 3}
library(msaviz)

aln_path <- system.file(
  "extdata", "hug_ribosomal.fa.gz", package = "msaviz"
)
aln <- seqinr::read.alignment(aln_path, format = "fasta")

alnDF <- msa2DF(aln)
msaHeatmap(alnDF)
```

By default `msaHeatmap()` colours every cell by whether it matches the
reference sequence (a computed consensus if you don't pass `reference =`).
Switch to per-letter colouring with `column = "Letter"`.

## Saving plots

`saveHeatmap()` is a `ggsave()` wrapper that picks a reasonable size based on
the number of sequences and axis text size:

```{r save, eval = FALSE}
saveHeatmap(msaHeatmap(alnDF), "aln.pdf")
```

## Shell script

A standalone CLI is shipped in `inst/scripts/msaHeatmap.R`. After installing
the package:

```{r cli-path, eval = FALSE}
system.file("scripts", "msaHeatmap.R", package = "msaviz")
```

Use it as:

```sh
Rscript msaHeatmap.R --input aln.fa --output aln.pdf --column Aln
```

See `--help` for all available options.

## Composing with PID, groups, and a dendrogram

`composeMSA()` lines the heatmap up with optional companion plots — a
percent-identity track, a hierarchical-clustering dendrogram, and a
left-side group annotation strip — using `patchwork`:

```{r compose, fig.width = 8, fig.height = 5}
d <- msaDendro(aln)
groups <- setNames(rep(c("Bacteria", "Archaea", "Eukarya"),
                       length.out = length(aln$nam)),
                   aln$nam)
group_cols <- c(Bacteria = "#1f77b4", Archaea = "#ff7f0e",
                Eukarya  = "#2ca02c")

composeMSA(
  heatmap = msaHeatmap(alnDF, row.order = attr(d, "order"),
                       legend.pos = "none"),
  top  = msaPID(alnDF),
  left = d,
  groups = groups[attr(d, "order")],
  group.colours = group_cols,
  heights = c(1, 6),       # PID short, heatmap tall
  widths  = c(0.5, 0.04, 4) # narrow dendrogram, thin group strip, wide heatmap
)
```

## More

The package vignette walks through `Aln` vs `Letter` colouring, custom
palettes, the `raster = TRUE/FALSE` trade-off, `posStats()`, and the
companion plotters:

```{r vignette, eval = FALSE}
vignette("msaviz")
```
