---
title: "Visualizing multiple sequence alignments with msaviz"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualizing multiple sequence alignments with msaviz}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 3.5,
  dpi = 120
)
```

`msaviz` turns a multiple sequence alignment into a tidy data frame and then
into a ggplot2 heatmap, with companion plots (percent-identity track,
dendrogram, group annotation) that compose cleanly via `patchwork`. This
vignette walks the full public API with a worked example for every knob.

```{r setup}
library(msaviz)
library(ggplot2)
```

# Loading an alignment

The package ships a 20-sequence × 250-position slice of the ribosomal-protein
alignment from Hug *et al.* (2016):

```{r load}
aln_path <- system.file("extdata", "hug_ribosomal.fa.gz", package = "msaviz")
aln <- seqinr::read.alignment(aln_path, format = "fasta")
length(aln$seq)
nchar(aln$seq[[1]])
```

`msaviz` doesn't depend on a particular alignment reader — anything that
yields a named character vector works. `Biostrings` multiple-alignment
classes are accepted too (gated here on the package being installed):

```{r biostrings}
if (requireNamespace("Biostrings", quietly = TRUE)) {
  bs_aln <- Biostrings::AAMultipleAlignment(c(
    s1 = "ACDEFGH-",
    s2 = "ACDEAGHK",
    s3 = "ACDEFGHK"
  ))
  msa2DF(bs_aln, reference = "s1")
}
```

# Reshaping with `msa2DF()`

`msa2DF()` reshapes the alignment into a tidy tibble with one row per
(sequence, position) cell. The `Aln` column records each cell's relationship
to the reference (`Ref` / `Alt` / `Gap`); the `Letter` column carries the
actual residue.

```{r msa2df-default}
alnDF <- msa2DF(aln)
alnDF
```

Because no `reference =` was passed, `msa2DF()` computed a per-position
consensus and used it as the reference. To compare against a specific
sequence instead, pass `reference =`:

```{r msa2df-ref}
ref_name <- aln$nam[1]
msa2DF(aln, reference = ref_name)
```

Other knobs you might reach for:

```{r msa2df-knobs, results = "hide"}
# Keep gap cells (Aln = "Gap", Letter = NA) instead of dropping them
msa2DF(aln, drop.gaps = FALSE)

# Keep the synthetic consensus row in the output
msa2DF(aln, keep.consensus = TRUE)

# Force everything to upper-case before comparing
msa2DF(aln, uppercase = TRUE)

# Ignore gaps when computing the consensus
msa2DF(aln, consensus.gaps = FALSE)
```

# Heatmaps with `msaHeatmap()`

The default heatmap colours each cell by whether it matches the reference,
which is the most useful view for closely related sequences:

```{r aln-default}
msaHeatmap(alnDF)
```

For per-residue colouring, set `column = "Letter"`. The package ships
sensible default palettes for the standard alphabets:

```{r letter-aa}
msaHeatmap(alnDF, column = "Letter", letter.colours = msa_palette_AA)
```

`msa_palette_DNA` / `msa_palette_RNA` cover the nucleotide alphabets:

```{r palettes-dna}
msa_palette_DNA

dna <- c(seq1 = "ACGTACGTACGT",
         seq2 = "ACGATCGTACGT",
         seq3 = "ACGTACGT-CGT",
         seq4 = "ACGTACGTNCGT")
msaHeatmap(msa2DF(dna, reference = "seq1"),
  column = "Letter", letter.colours = msa_palette_DNA, raster = FALSE)
```

Or define your own:

```{r letter-custom}
spectral_palette <- setNames(
  grDevices::hcl.colors(nlevels(alnDF$Letter), palette = "Spectral"),
  levels(alnDF$Letter)
)
msaHeatmap(alnDF, column = "Letter", letter.colours = spectral_palette)
```

## Layout knobs

Move the row labels, the legend, and trim long names:

```{r layout-names}
msaHeatmap(alnDF, names.pos = "left", legend.pos = "bottom",
  trim.names = TRUE, trim.names.nchar = 30)
```

Drop the legend entirely and hide axis labels for a clean panel:

```{r layout-bare}
msaHeatmap(alnDF, legend.pos = "none", hide.x.labels = TRUE,
  hide.y.labels = TRUE)
```

Adjust text sizes axis by axis:

```{r layout-text}
msaHeatmap(alnDF, text.size.x = 5, text.size.y = 4, text.size.legend = 7)
```

Override the x-axis breaks. The package's [`integer_breaks()`][integer_breaks]
helper is what `msaHeatmap()` uses internally:

```{r layout-breaks}
integer_breaks(c(1, 250))
msaHeatmap(alnDF, x.breaks = c(1, 50, 100, 150, 200, 250),
  x.labels = paste0("p", c(1, 50, 100, 150, 200, 250)))
```

## Custom row order

`row.order =` controls top-to-bottom sequence order:

```{r row-order}
custom <- rev(aln$nam[1:8])  # take the first 8, plot bottom-to-top
sub_df <- subset(alnDF, Sequence %in% custom)
msaHeatmap(sub_df, row.order = custom)
```

## Stacking ggplot2 layers on top

`msaHeatmap()` returns a regular ggplot object, so any `+ geom_*()` or
`+ theme()` works:

```{r stack-layers}
msaHeatmap(alnDF, legend.pos = "none", hide.y.labels = TRUE) +
  annotate("rect", xmin = 80, xmax = 120, ymin = 0.5, ymax = 20.5,
    fill = NA, colour = "blue", linewidth = 0.4) +
  labs(title = "Highlighting positions 80-120")
```

# Saving with `saveHeatmap()`

`saveHeatmap()` is a `ggsave()` wrapper that picks a reasonable size from
the number of sequences and the axis text size:

```{r save-auto, eval = FALSE}
saveHeatmap(msaHeatmap(alnDF), tempfile(fileext = ".pdf"))
```

Pass `height =` / `width =` to override:

```{r save-fixed, eval = FALSE}
saveHeatmap(msaHeatmap(alnDF), tempfile(fileext = ".png"),
  height = 70, width = 200, unit = "mm")
```

# Rasterization

Alignments with hundreds of thousands of cells produce huge SVG/PDF output
when drawn as vector rectangles. `msaHeatmap(raster = TRUE)` (the default)
sends the tile layer through `rasterGeom()` — drawn into an in-memory PNG
at `raster.dpi` and laid back onto the panel as a single bitmap.

```{r raster-default}
msaHeatmap(alnDF, raster = TRUE, raster.dpi = 96, legend.pos = "none")
```

Turn rasterization off for crisp vector output on small alignments:

```{r raster-off}
msaHeatmap(alnDF, raster = FALSE, legend.pos = "none")
```

`rasterGeom()` works outside of `msaHeatmap()` too. Drop it into any
ggplot2 layer that benefits from rasterization:

```{r rastergeom-standalone}
set.seed(1)
dots <- data.frame(x = runif(5000), y = runif(5000))
ggplot(dots, aes(x, y)) +
  rasterGeom(geom_point(size = 0.6, colour = "steelblue"), png.dpi = 96) +
  theme_minimal()
```

# Position summaries with `posStats()`

`posStats()` returns one row per alignment position — counts, gap counts,
the fraction of sequences matching the reference, the most common letter
and its frequency, plus Shannon entropy:

```{r posstats}
ps <- posStats(alnDF)
ps
```

By default gaps are excluded from `mostCommon`/`entropy`. Pass
`include.gaps = TRUE` to count them as just another letter:

```{r posstats-gaps}
posStats(msa2DF(aln, drop.gaps = FALSE), include.gaps = TRUE)
```

Use the tibble directly with ggplot2 for ad-hoc per-position plots:

```{r posstats-entropy, fig.height = 2.4}
ggplot(ps, aes(Position, entropy)) +
  geom_col(fill = "grey40") +
  scale_x_continuous(breaks = integer_breaks(c(1, max(ps$Position))),
    name = NULL, expand = c(0, 0)) +
  labs(y = "Shannon entropy", title = "Per-position entropy") +
  theme_minimal()
```

```{r posstats-gaps-bar, fig.height = 2.4}
ggplot(ps, aes(Position, nGaps)) +
  geom_col(fill = "tomato") +
  scale_x_continuous(breaks = integer_breaks(c(1, max(ps$Position))),
    name = NULL, expand = c(0, 0)) +
  labs(y = "# gap sequences", title = "Gappy positions") +
  theme_minimal()
```

# Consensus with `consensusSeq()`

`consensusSeq()` is the same per-position consensus calculation that
`msa2DF()` performs internally, exposed as a standalone function:

```{r consensus}
substr(consensusSeq(aln), 1, 60)
```

With `include.gaps = FALSE`, gaps are excluded when picking the most
common letter at each position:

```{r consensus-nogaps}
substr(consensusSeq(aln, include.gaps = FALSE), 1, 60)
```

# Percent-identity track with `msaPID()`

`msaPID()` draws a per-position % identity track sized to line up with
`msaHeatmap()`'s x-axis, ready to be stacked above it.

Default style is bars. The y-axis sits on the right by default, matching
`msaHeatmap()`'s default `names.pos = "right"` so the panels' left edges
align cleanly when composed:

```{r pid-bar, fig.height = 2}
msaPID(alnDF)
```

Switch to a step line:

```{r pid-line, fig.height = 2}
msaPID(alnDF, style = "line", line.colour = "steelblue")
```

For a standalone PID plot (without a heatmap below it), the conventional
left-side y-axis reads more naturally:

```{r pid-standalone, fig.height = 2.4}
msaPID(alnDF, fill = "darkorchid4", hide.x.labels = FALSE,
  y.axis.pos = "left")
```

# Dendrogram with `msaDendro()`

`msaDendro()` computes pairwise distances with `seqinr::dist.alignment()`,
clusters with `hclust()`, and returns a ggplot dendrogram whose y-axis
lines up with the heatmap's row order. The leaf order is attached as the
`"order"` attribute.

Default orientation — root on the left, leaves on the right (drop-in
companion for a heatmap):

```{r dendro-flipped, fig.width = 3, fig.height = 5}
d <- msaDendro(aln)
d
head(attr(d, "order"))
```

Classical bottom-up orientation:

```{r dendro-classical, fig.width = 5, fig.height = 3}
msaDendro(aln, flip = FALSE)
```

Other distance / linkage choices:

```{r dendro-alt, fig.width = 3, fig.height = 5}
msaDendro(aln, dist.matrix = "similarity", hclust.method = "complete")
```

# Composing with `composeMSA()`

`composeMSA()` lays out a heatmap with optional companion plots using
`patchwork` under the hood. Each of the four roles — heatmap, top, left,
groups — is independent.

## Heatmap + PID

The fill legend takes up vertical space between the top track and the
heatmap, so it's usually cleaner to suppress it with
`legend.pos = "none"` whenever a `top` companion is present:

```{r compose-pid, fig.height = 4.5}
composeMSA(
  heatmap = msaHeatmap(alnDF, legend.pos = "none"),
  top     = msaPID(alnDF)
)
```

## Heatmap + dendrogram

When `left` is an `msaDendro()` plot, `composeMSA()` reads its `"order"`
attribute and warns if the heatmap was built with a different row order:

```{r compose-dendro, fig.height = 4.5}
hm <- msaHeatmap(alnDF, row.order = attr(d, "order"))
composeMSA(heatmap = hm, left = d)
```

## Heatmap + group strip

A toy grouping (in practice you'd pull these labels from your metadata):

```{r compose-groups, fig.height = 4}
groups <- setNames(
  rep(c("Bacteria", "Archaea", "Eukarya"), length.out = length(aln$nam)),
  aln$nam
)
group_cols <- c(Bacteria = "#1f77b4", Archaea = "#ff7f0e",
                Eukarya = "#2ca02c")
composeMSA(
  heatmap = msaHeatmap(alnDF),
  groups = groups,
  group.colours = group_cols
)
```

## Everything together

```{r compose-all, fig.height = 5}
composeMSA(
  heatmap = msaHeatmap(alnDF, row.order = attr(d, "order"),
                       legend.pos = "none"),
  top  = msaPID(alnDF),
  left = d,
  groups = groups[attr(d, "order")],
  group.colours = group_cols
)
```

Adjust the relative panel sizes with `heights =` and `widths =`:

```{r compose-sizes, fig.height = 4.5}
composeMSA(
  heatmap = msaHeatmap(alnDF, row.order = attr(d, "order"),
                       legend.pos = "none"),
  top  = msaPID(alnDF),
  left = d,
  heights = c(1, 6),       # PID gets 1, heatmap gets 6
  widths  = c(0.5, 4)      # dendrogram narrow, heatmap wide
)
```

## Custom compositions

Because every piece is a regular ggplot (or patchwork), you can compose
manually too. For a side-by-side comparison of `Aln` vs `Letter`
colourings sharing the same y-axis:

```{r custom-side-by-side, fig.height = 4.5}
patchwork::wrap_plots(
  msaHeatmap(alnDF, legend.pos = "none"),
  msaHeatmap(alnDF, column = "Letter", letter.colours = msa_palette_AA,
    legend.pos = "none", hide.y.labels = TRUE),
  nrow = 1
)
```

# Shell script

A standalone CLI ships in `inst/scripts/msaHeatmap.R`:

```{r cli-path, eval = FALSE}
system.file("scripts", "msaHeatmap.R", package = "msaviz")
```

Use it as:

```sh
Rscript msaHeatmap.R --input aln.fa --output aln.pdf --column Aln \
  --reference my_ref --keep-gaps --width 200 --height 70
```

See `--help` for all available flags.

# See also

* [`msa2DF()`][msa2DF] — the reshaping step. `reference =`, `drop.gaps =`,
  `keep.consensus =`, `consensus.gaps =`, `uppercase =`, `verbose =`.
* [`msaHeatmap()`][msaHeatmap] — the main plot.
* [`saveHeatmap()`][saveHeatmap] — auto-sized `ggsave()` wrapper.
* [`rasterGeom()`][rasterGeom] — usable outside `msaHeatmap()` to
  rasterize any ggplot geom.
* [`consensusSeq()`][consensusSeq] — standalone consensus.
* [`posStats()`][posStats] — per-position summary tibble.
* [`msaPID()`][msaPID] — percent-identity track.
* [`msaDendro()`][msaDendro] — clustering dendrogram.
* [`composeMSA()`][composeMSA] — patchwork composition.
* [`integer_breaks()`][integer_breaks] — integer-only tick positions.

[msa2DF]: ../reference/msa2DF.html
[msaHeatmap]: ../reference/msaHeatmap.html
[saveHeatmap]: ../reference/saveHeatmap.html
[rasterGeom]: ../reference/rasterGeom.html
[consensusSeq]: ../reference/consensusSeq.html
[posStats]: ../reference/posStats.html
[msaPID]: ../reference/msaPID.html
[msaDendro]: ../reference/msaDendro.html
[composeMSA]: ../reference/composeMSA.html
[integer_breaks]: ../reference/integer_breaks.html
