---
title: "Visualizing multiple sequence alignments with msaviz"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualizing multiple sequence alignments with msaviz}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 3.5,
  dpi = 120
)
```

`msaviz` turns a multiple sequence alignment into a tidy data frame and then
into a ggplot2 heatmap, with a handful of companion plots (a percent-identity
track, a dendrogram, group annotation) that compose cleanly via `patchwork`.
This vignette walks through the full public API, with a worked example for
each knob along the way.

```{r setup}
library(msaviz)
library(ggplot2)
```

# Loading an alignment

The package ships a 20-sequence × 250-position slice of the ribosomal-protein
alignment from Hug *et al.* (2016):

```{r load}
aln_path <- system.file("extdata", "hug_ribosomal.fa.gz", package = "msaviz")
aln <- seqinr::read.alignment(aln_path, format = "fasta")
length(aln$seq)
nchar(aln$seq[[1]])
```

`msaviz` doesn't depend on a particular alignment reader; anything that
yields a named character vector will work. `Biostrings` multiple-alignment
classes are accepted too (gated here on the package actually being
installed):

```{r biostrings}
if (requireNamespace("Biostrings", quietly = TRUE)) {
  bs_aln <- Biostrings::AAMultipleAlignment(c(
    s1 = "ACDEFGH-",
    s2 = "ACDEAGHK",
    s3 = "ACDEFGHK"
  ))
  msa2DF(bs_aln, reference = "s1")
}
```

# Reshaping with `msa2DF()`

`msa2DF()` reshapes the alignment into a tidy tibble with one row per
(sequence, position) cell. The `Aln` column records each cell's relationship
to the reference (`Ref` / `Alt` / `Gap`), and the `Letter` column carries the
actual residue.

```{r msa2df-default}
alnDF <- msa2DF(aln)
alnDF
```

(Note that the gap cells and the synthetic consensus row are both dropped by
default, so the tibble has fewer rows than the full sequence × position
grid.)

Because no `reference =` was passed, `msa2DF()` computed a per-position
consensus and used that as the reference. To compare against a specific
sequence instead, just pass `reference =`:

```{r msa2df-ref}
ref_name <- aln$nam[1]
msa2DF(aln, reference = ref_name)
```

Other knobs you might reach for:

```{r msa2df-knobs, results = "hide"}
# Keep gap cells (Aln = "Gap", Letter = NA) instead of dropping them
msa2DF(aln, drop.gaps = FALSE)

# Keep the synthetic consensus row in the output
msa2DF(aln, keep.consensus = TRUE)

# Force everything to upper-case before comparing
msa2DF(aln, uppercase = TRUE)

# Ignore gaps when computing the consensus
msa2DF(aln, consensus.gaps = FALSE)
```

# Heatmaps with `msaHeatmap()`

The default heatmap colours each cell by whether it matches the reference,
which is often the most useful view for closely related sequences:

```{r aln-default}
msaHeatmap(alnDF)
```

For per-residue colouring, set `column = "Letter"`. The package ships with
sensible default palettes for the standard alphabets, so you don't have to
build one yourself:

```{r letter-aa}
msaHeatmap(alnDF, column = "Letter", letter.colours = msa_palette_AA)
```

`msa_palette_DNA` / `msa_palette_RNA` cover the nucleotide alphabets:

```{r palettes-dna}
msa_palette_DNA

dna <- c(seq1 = "ACGTACGTACGT",
         seq2 = "ACGATCGTACGT",
         seq3 = "ACGTACGT-CGT",
         seq4 = "ACGTACGTNCGT")
msaHeatmap(msa2DF(dna, reference = "seq1"),
  column = "Letter", letter.colours = msa_palette_DNA, raster = FALSE)
```

Or define your own:

```{r letter-custom}
spectral_palette <- setNames(
  grDevices::hcl.colors(nlevels(alnDF$Letter), palette = "Spectral"),
  levels(alnDF$Letter)
)
msaHeatmap(alnDF, column = "Letter", letter.colours = spectral_palette)
```

## Layout knobs

Move the row labels, the legend, and trim long names:

```{r layout-names}
msaHeatmap(alnDF, names.pos = "left", legend.pos = "bottom",
  trim.names = TRUE, trim.names.nchar = 30)
```

Drop the legend entirely and hide axis labels for a clean panel:

```{r layout-bare}
msaHeatmap(alnDF, legend.pos = "none", hide.x.labels = TRUE,
  hide.y.labels = TRUE)
```

Adjust text sizes axis by axis:

```{r layout-text}
msaHeatmap(alnDF, text.size.x = 5, text.size.y = 4, text.size.legend = 7)
```

Override the x-axis breaks. The package's [`integer_breaks()`][integer_breaks]
helper is what `msaHeatmap()` uses internally:

```{r layout-breaks}
integer_breaks(c(1, 250))
msaHeatmap(alnDF, x.breaks = c(1, 50, 100, 150, 200, 250),
  x.labels = paste0("p", c(1, 50, 100, 150, 200, 250)))
```

To annotate every position on a long alignment, combine a denser `x.breaks =`
with `x.labels.rotate =` so the labels fit vertically:

```{r layout-rotate}
msaHeatmap(alnDF,
  x.breaks = seq(1, 250, by = 10),
  x.labels.rotate = 90,
  text.size.x = 4)
```

`y.labels.rotate =` does the same for the sequence-name axis.

## Custom row order

`row.order =` controls top-to-bottom sequence order:

```{r row-order}
custom <- rev(aln$nam[1:8])  # take the first 8, plot bottom-to-top
sub_df <- subset(alnDF, Sequence %in% custom)
msaHeatmap(sub_df, row.order = custom)
```

## Stacking ggplot2 layers on top

`msaHeatmap()` returns a regular ggplot object, so any `+ geom_*()` or
`+ theme()` works:

```{r stack-layers}
msaHeatmap(alnDF, legend.pos = "none", hide.y.labels = TRUE) +
  annotate("rect", xmin = 80, xmax = 120, ymin = 0.5, ymax = 20.5,
    fill = NA, colour = "blue", linewidth = 0.4) +
  labs(title = "Highlighting positions 80-120")
```

## Emphasising specific cells

In a dense heatmap a single non-reference cell can end up only one pixel
wide, which is easy to miss when zoomed out. `emphasize.size` adds a second
tile layer on top of the base layer, with enlarged tiles, so that the cells
of interest overlap their neighbours and stay visible at any zoom level.

The defaults (`emphasize = "Alt"`, `emphasize.by = "Aln"`) already match the
"make the SNPs pop" case, so bumping `emphasize.size` on its own is enough:

```{r emphasize-snp}
msaHeatmap(alnDF, emphasize.size = 1.5)
```

`emphasize.size.x` and `emphasize.size.y` allow asymmetric expansion: a
wider-than-tall overlay reads as a vertical streak in long alignments,
making column-conserved differences pop:

```{r emphasize-asym}
msaHeatmap(alnDF, emphasize.size.x = 2.5, emphasize.size.y = 1.2)
```

The feature is general: `emphasize.by =` picks which column of `alnDF` to
match against, and `emphasize =` lists the values to enlarge. So to
emphasise specific residues in a `column = "Letter"` heatmap (say,
highlighting the cysteines and tryptophans):

```{r emphasize-letter}
msaHeatmap(alnDF,
  column = "Letter", letter.colours = msa_palette_AA,
  emphasize = c("C", "W"),
  emphasize.by = "Letter",
  emphasize.size = 1.8)
```

Set `emphasize.size = 1` (the default) to disable the overlay entirely.

# Saving with `saveHeatmap()`

`saveHeatmap()` is a `ggsave()` wrapper that picks a reasonable size from
the number of sequences and the axis text size:

```{r save-auto, eval = FALSE}
saveHeatmap(msaHeatmap(alnDF), tempfile(fileext = ".pdf"))
```

Pass `height =` / `width =` to override:

```{r save-fixed, eval = FALSE}
saveHeatmap(msaHeatmap(alnDF), tempfile(fileext = ".png"),
  height = 70, width = 200, unit = "mm")
```

# Rasterization

Alignments with hundreds of thousands of cells produce enormous SVG/PDF
output when drawn as vector rectangles (one rectangle per cell adds up
quickly). `msaHeatmap(raster = TRUE)`, which is the default, sends the tile
layer through `rasterGeom()`, which draws it into an in-memory PNG at
`raster.dpi` and lays it back onto the panel as a single bitmap.

```{r raster-default}
msaHeatmap(alnDF, raster = TRUE, raster.dpi = 96, legend.pos = "none")
```

Turn rasterization off for crisp vector output on small alignments:

```{r raster-off}
msaHeatmap(alnDF, raster = FALSE, legend.pos = "none")
```

`rasterGeom()` works outside of `msaHeatmap()` too. Drop it into any
ggplot2 layer that benefits from rasterization:

```{r rastergeom-standalone}
set.seed(1)
dots <- data.frame(x = runif(5000), y = runif(5000))
ggplot(dots, aes(x, y)) +
  rasterGeom(geom_point(size = 0.6, colour = "steelblue"), png.dpi = 96) +
  theme_minimal()
```

# Position summaries with `posStats()`

`posStats()` returns one row per alignment position with counts, gap
counts, the fraction of sequences matching the reference, the most
common letter and its frequency, plus Shannon entropy:

```{r posstats}
ps <- posStats(alnDF)
ps
```

By default gaps are excluded from `mostCommon`/`entropy`. Pass
`include.gaps = TRUE` to count them as just another letter:

```{r posstats-gaps}
posStats(msa2DF(aln, drop.gaps = FALSE), include.gaps = TRUE)
```

Use the tibble directly with ggplot2 for ad-hoc per-position plots:

```{r posstats-entropy, fig.height = 2.4}
ggplot(ps, aes(Position, entropy)) +
  geom_col(fill = "grey40") +
  scale_x_continuous(breaks = integer_breaks(c(1, max(ps$Position))),
    name = NULL, expand = c(0, 0)) +
  labs(y = "Shannon entropy", title = "Per-position entropy") +
  theme_minimal()
```

```{r posstats-gaps-bar, fig.height = 2.4}
ggplot(ps, aes(Position, nGaps)) +
  geom_col(fill = "tomato") +
  scale_x_continuous(breaks = integer_breaks(c(1, max(ps$Position))),
    name = NULL, expand = c(0, 0)) +
  labs(y = "# gap sequences", title = "Gappy positions") +
  theme_minimal()
```

# Consensus with `consensusSeq()`

`consensusSeq()` is the same per-position consensus calculation that
`msa2DF()` performs internally, exposed as a standalone function:

```{r consensus}
substr(consensusSeq(aln), 1, 60)
```

With `include.gaps = FALSE`, gaps are excluded when picking the most
common letter at each position:

```{r consensus-nogaps}
substr(consensusSeq(aln, include.gaps = FALSE), 1, 60)
```

# Percent-identity track with `msaPID()`

`msaPID()` draws a per-position % identity track sized to line up with
`msaHeatmap()`'s x-axis, ready to be stacked above it.

Default style is bars. The y-axis sits on the right by default, matching
`msaHeatmap()`'s default `names.pos = "right"` so the panels' left edges
align cleanly when composed:

```{r pid-bar, fig.height = 2}
msaPID(alnDF)
```

Switch to a step line:

```{r pid-line, fig.height = 2}
msaPID(alnDF, style = "line", line.colour = "steelblue")
```

For a standalone PID plot (without a heatmap below it), the conventional
left-side y-axis reads more naturally:

```{r pid-standalone, fig.height = 2.4}
msaPID(alnDF, fill = "darkorchid4", hide.x.labels = FALSE,
  y.axis.pos = "left")
```

# Dendrogram with `msaDendro()`

`msaDendro()` computes pairwise distances with `seqinr::dist.alignment()`,
clusters with `hclust()`, and returns a ggplot dendrogram whose y-axis
lines up with the heatmap's row order. The leaf order is attached as the
`"order"` attribute.

Default orientation, with the root on the left and the leaves on the
right (drop-in companion for a heatmap):

```{r dendro-flipped, fig.width = 3, fig.height = 5}
d <- msaDendro(aln)
d
head(attr(d, "order"))
```

Classical bottom-up orientation:

```{r dendro-classical, fig.width = 5, fig.height = 3}
msaDendro(aln, flip = FALSE)
```

Other distance / linkage choices:

```{r dendro-alt, fig.width = 3, fig.height = 5}
msaDendro(aln, dist.matrix = "similarity", hclust.method = "complete")
```

# Per-sequence lollipops with `msaLollipop()`

`msaLollipop()` is an alternative view of the same `alnDF`. One row of
a discrete y-axis is allocated per sequence, with each non-reference
cell drawn as a vertical stem topped with a circular head and an
optional `RefLetterPositionAltLetter` label. A per-position horizontal
baseline at each row makes gaps in the alignment show up as visible
breaks. This is the layout used for Supplementary Figure 20 of Hodgins
et al. (2023).

Because `msaLollipop()` uses a single discrete y-axis (like
`msaHeatmap()`), it composes cleanly with `composeMSA()`. With labels
on, lollipops are most readable on small alignments or short windows
of larger ones, so we use a toy 4-sequence alignment to introduce
the layout. Build the `alnDF` with `keep.consensus = TRUE` so the
`RefLetterPositionAltLetter` labels can include the reference letter:

```{r lollipop-default, fig.height = 2.4}
toy <- c(
  seq1 = "ACGTACGTACGTACGTACGT",
  seq2 = "ACGTACGAACGTACGTACGT",
  seq3 = "AAGTACGTACGTGCGTACGT",
  seq4 = "ACGTACGTACGTACGTCCGA"
)
toyDF <- msa2DF(toy, reference = "seq1", drop.gaps = FALSE,
  keep.consensus = TRUE)
msaLollipop(toyDF)
```

Head fill can be mapped to any column of `alnDF` via `head.fill.by =`
with an optional palette in `head.fill.colours =`. Colour by alt
residue:

```{r lollipop-fill, fig.height = 2.4}
msaLollipop(toyDF,
  head.fill.by = "Letter",
  head.fill.colours = msa_palette_DNA)
```

On a dense alignment, labels overlap, so the unlabeled form
reproduces the paper's "lollipop_snps2_nolab" variant on the full
20-sequence ribosomal slice:

```{r lollipop-no-labels, fig.height = 5}
alnDF_ll <- msa2DF(aln, drop.gaps = FALSE, keep.consensus = TRUE)
msaLollipop(alnDF_ll, labels = FALSE)
```

`row.order =` accepts names that are not present in the data; they
render as empty rows, reproducing the paper's "REF spacer" trick for
separating groups of sequences. Shown here on the toy alignment so
the spacers are visible at a glance:

```{r lollipop-spacers, fig.height = 3.5}
custom_order <- c("seq1", "spacer1", "seq2", "spacer2", "seq3", "seq4")
msaLollipop(toyDF, drop.empty = FALSE, row.order = custom_order)
```

Stem heights are constant by default. Pass a column name to `y =` and
the stems scale with that column. A synthetic example (a random
per-cell score) on the full alignment:

```{r lollipop-y, fig.height = 5}
set.seed(1)
alnDF_ll$score <- runif(nrow(alnDF_ll), 0.2, 1)
msaLollipop(alnDF_ll, y = "score", labels = FALSE)
```

`msaLollipop()` is a drop-in substitute for `msaHeatmap()` inside
`composeMSA()`; the composition example sits in the next section.

# Composing with `composeMSA()`

`composeMSA()` lays out a heatmap with optional companion plots using
`patchwork` under the hood. Each of the four roles (heatmap, top, left,
groups) is independent.

## Heatmap + PID

The fill legend takes up vertical space between the top track and the
heatmap, so it's usually cleaner to suppress it with
`legend.pos = "none"` whenever a `top` companion is present:

```{r compose-pid, fig.height = 4.5}
composeMSA(
  heatmap = msaHeatmap(alnDF, legend.pos = "none"),
  top     = msaPID(alnDF)
)
```

## Heatmap + dendrogram

When `left` is an `msaDendro()` plot, `composeMSA()` reads its `"order"`
attribute and warns if the heatmap was built with a different row order:

```{r compose-dendro, fig.height = 4.5}
hm <- msaHeatmap(alnDF, row.order = attr(d, "order"))
composeMSA(heatmap = hm, left = d)
```

## Heatmap + group strip

A toy grouping (in practice you'd pull these labels from your metadata):

```{r compose-groups, fig.height = 4}
groups <- setNames(
  rep(c("Bacteria", "Archaea", "Eukarya"), length.out = length(aln$nam)),
  aln$nam
)
group_cols <- c(Bacteria = "#1f77b4", Archaea = "#ff7f0e",
                Eukarya = "#2ca02c")
composeMSA(
  heatmap = msaHeatmap(alnDF),
  groups = groups,
  group.colours = group_cols
)
```

## Heatmap + column group strip

`col.groups =` is the same idea on the x-axis: a length-`aln.size` vector
where each entry labels one alignment position. Adjacent identical labels
render as a single contiguous block, so runs of positions (UTR / CDS /
UTR, exons, domains, primer regions, ...) are easy to build with `rep()`:

```{r compose-col-groups, fig.height = 4}
aln_size <- attr(alnDF, "aln.size")
col_groups <- rep(c("5'UTR", "CDS", "3'UTR"),
  c(60, 120, aln_size - 180))
col_cols <- c("5'UTR" = "#a6cee3", "CDS" = "#1f78b4",
              "3'UTR" = "#b2df8a")
composeMSA(
  heatmap = msaHeatmap(alnDF, legend.pos = "none"),
  col.groups = col_groups,
  col.group.colours = col_cols
)
```

Row and column strips compose together:

```{r compose-both-strips, fig.height = 4}
composeMSA(
  heatmap = msaHeatmap(alnDF, legend.pos = "none"),
  groups = groups,
  group.colours = group_cols,
  col.groups = col_groups,
  col.group.colours = col_cols
)
```

## Everything together

```{r compose-all, fig.height = 5}
composeMSA(
  heatmap = msaHeatmap(alnDF, row.order = attr(d, "order"),
                       legend.pos = "none"),
  top  = msaPID(alnDF),
  left = d,
  groups = groups[attr(d, "order")],
  group.colours = group_cols
)
```

Adjust the relative panel sizes with `heights =` and `widths =`:

```{r compose-sizes, fig.height = 4.5}
composeMSA(
  heatmap = msaHeatmap(alnDF, row.order = attr(d, "order"),
                       legend.pos = "none"),
  top  = msaPID(alnDF),
  left = d,
  heights = c(1, 6),       # PID gets 1, heatmap gets 6
  widths  = c(0.5, 4)      # dendrogram narrow, heatmap wide
)
```

## Lollipop in place of the heatmap

Because `msaLollipop()` carries the same `row.order` and `aln.size`
attributes as `msaHeatmap()`, it can be passed straight through as the
`heatmap =` argument and every companion plot still lines up:

```{r compose-lollipop, fig.height = 6}
composeMSA(
  heatmap = msaLollipop(alnDF_ll, row.order = attr(d, "order"),
    labels = FALSE),
  top  = msaPID(alnDF_ll),
  left = d
)
```

## Custom compositions

Because every piece is a regular ggplot (or patchwork), you can compose
manually too. For a side-by-side comparison of `Aln` vs `Letter`
colourings sharing the same y-axis:

```{r custom-side-by-side, fig.height = 4.5}
patchwork::wrap_plots(
  msaHeatmap(alnDF, legend.pos = "none"),
  msaHeatmap(alnDF, column = "Letter", letter.colours = msa_palette_AA,
    legend.pos = "none", hide.y.labels = TRUE),
  nrow = 1
)
```

# Shell scripts

Two standalone command-line scripts ship with the package:
`inst/scripts/msaHeatmap.R` for the heatmap view, and
`inst/scripts/msaLollipop.R` for the lollipop view.

```{r cli-path, eval = FALSE}
system.file("scripts", "msaHeatmap.R",  package = "msaviz")
system.file("scripts", "msaLollipop.R", package = "msaviz")
```

Use them as:

```sh
Rscript msaHeatmap.R  --input aln.fa --output aln.pdf      --column Aln \
  --reference my_ref --keep-gaps --width 200 --height 70

Rscript msaLollipop.R --input aln.fa --output lollipop.pdf \
  --reference my_ref --width 11.5 --height 7
```

See `--help` on each for all available flags.

# See also

* [`msa2DF()`][msa2DF]: the reshaping step. `reference =`, `drop.gaps =`,
  `keep.consensus =`, `consensus.gaps =`, `uppercase =`, `verbose =`.
* [`msaHeatmap()`][msaHeatmap]: the main plot.
* [`saveHeatmap()`][saveHeatmap]: auto-sized `ggsave()` wrapper.
* [`rasterGeom()`][rasterGeom]: usable outside `msaHeatmap()` to
  rasterize any ggplot geom.
* [`consensusSeq()`][consensusSeq]: standalone consensus.
* [`posStats()`][posStats]: per-position summary tibble.
* [`msaPID()`][msaPID]: percent-identity track.
* [`msaDendro()`][msaDendro]: clustering dendrogram.
* [`msaLollipop()`][msaLollipop]: per-sequence lollipop chart.
* [`composeMSA()`][composeMSA]: patchwork composition.
* [`integer_breaks()`][integer_breaks]: integer-only tick positions.

[msa2DF]: ../reference/msa2DF.html
[msaHeatmap]: ../reference/msaHeatmap.html
[saveHeatmap]: ../reference/saveHeatmap.html
[rasterGeom]: ../reference/rasterGeom.html
[consensusSeq]: ../reference/consensusSeq.html
[posStats]: ../reference/posStats.html
[msaPID]: ../reference/msaPID.html
[msaDendro]: ../reference/msaDendro.html
[msaLollipop]: ../reference/msaLollipop.html
[composeMSA]: ../reference/composeMSA.html
[integer_breaks]: ../reference/integer_breaks.html
