---
title: "Visualizing multiple sequence alignments with msaviz"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualizing multiple sequence alignments with msaviz}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 3.5,
  dpi = 120
)
```

`msaviz` turns a multiple sequence alignment into a tidy data frame and then
into a ggplot2 heatmap. This vignette walks through the bundled example
alignment and shows the main knobs you'll want to reach for.

```{r setup}
library(msaviz)
library(ggplot2)
```

## A small example alignment

The package ships a 20-sequence × 250-position slice of the ribosomal-protein
alignment from Hug *et al.* (2016):

```{r load}
aln_path <- system.file("extdata", "hug_ribosomal.fa.gz", package = "msaviz")
aln <- seqinr::read.alignment(aln_path, format = "fasta")
alnDF <- msa2DF(aln)
alnDF
```

`msa2DF()` returns one row per (sequence, position) cell. The `Aln` column
records whether each cell matches the reference (`Ref`), differs from it
(`Alt`), or is a gap (`Gap`); the `Letter` column carries the actual amino-
acid character.

Because we didn't pass a `reference =`, `msa2DF()` built a consensus sequence
and compared every other sequence against it.

## Reference- vs letter-coloured heatmaps

The default `column = "Aln"` colouring highlights deviations from the
reference, which is usually the most informative view for closely related
sequences:

```{r aln-default}
msaHeatmap(alnDF)
```

For a per-residue view, use `column = "Letter"`. The default palette is the
ggplot2 discrete scale; pass `letter.colours =` to supply your own:

```{r letter}
msaHeatmap(alnDF, column = "Letter")
```

```{r letter-custom}
aa_palette <- setNames(
  grDevices::hcl.colors(nlevels(alnDF$Letter), palette = "Spectral"),
  levels(alnDF$Letter)
)
msaHeatmap(alnDF, column = "Letter", letter.colours = aa_palette)
```

## Rasterizing for speed and small files

Large heatmaps render slowly and produce huge vector files. The `raster =
TRUE` default sends the tile layer through [`rasterGeom()`][rasterGeom] —
that draws the tiles into an in-memory PNG and then lays the bitmap back
onto the panel. Output sizes drop from megabytes to kilobytes, and screen
rendering becomes instant.

To compare, build a non-rasterized version and time it:

```{r raster-off, eval = FALSE}
msaHeatmap(alnDF, raster = FALSE)
```

You'd normally only set `raster = FALSE` for very small alignments where you
want crisp vector output.

## Saving to disk

`saveHeatmap()` is a `ggsave()` wrapper. When you don't pass a size, it
picks a height that grows with the number of sequences and a width at least
twice that height (or 100 mm, whichever is bigger):

```{r save, eval = FALSE}
saveHeatmap(msaHeatmap(alnDF), tempfile(fileext = ".pdf"))
```

## Extending the plot

`msaHeatmap()` returns a regular `ggplot` object, so anything you'd normally
add with `+` works:

```{r facet}
msaHeatmap(alnDF) +
  theme(axis.text.y = element_text(size = 5)) +
  labs(title = "Hug et al. (2016) ribosomal proteins — example slice")
```

## See also

* [`msa2DF()`][msa2DF] — the reshaping step. Use `reference =` to compare
  against a specific sequence rather than the consensus, and
  `drop.gaps = FALSE` to keep gap rows for downstream filtering.
* [`rasterGeom()`][rasterGeom] — usable outside of `msaHeatmap()` to
  rasterize any ggplot geom.

[msa2DF]: ../reference/msa2DF.html
[msaHeatmap]: ../reference/msaHeatmap.html
[rasterGeom]: ../reference/rasterGeom.html
